Measuring the Inherent Readability of Five Common Programming Languages

### 2015-03-17 rmtw/a1/assignment_1.tex
Updated assignment 1 based on feedback

%
%  $Description: Author guidelines and sample document in LaTeX 2.09$ 
%
%  $Author: ienne $
%  $Date: 1995/09/15 15:20:59 $
%  $Revision: 1.4 $
%

\documentclass[times, 10pt,twocolumn]{Article} 
\usepackage{latex8}
\usepackage{times}
\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage{tabularx}

@@

%------------------------------------------------------------------------- 
\Section{Introduction}
%Problem and definition
Comparisons between programming languages are difficult to perform, due to inherent differences between languages, such as syntax. One aspect of code is its readability, which in turn is a determining factor of a piece of software's quality and maintainability \cite{aggarwal2002integrated, elshoff1982improving}. Buse \& Westley define readability as ``a property that influences how easily a given piece of code can be read and understood'' \cite{buse2010learning}.

@@

Industry and academia both face the problem of code maintainability and by association, readability. As was highlighted previously, the readability of code can affect its maintainability. This study could help some in industry to decide which languages should be used if maintainability is of high concern. Further studies could be conducted by to determine if there is a correlation between the popularity of projects and the readability of their languages. Furthermore, academia would be interested in which languages are more popular and those that have a higher readability, especially for novice developers. Further investigation into the design of these programming languages could yield insights into what makes them so popular and potentially well suited for teaching.

The purpose of this survey study is therefore to test what are the differences in readability between five common and well known programming languages. With the aim of suggesting a metric which describes the readability of a given program, taking into account the inherent readability of the language in which it is written.

@@

%------------------------------------------------------------------------- 
\SubSection{Research Questions}
This study has been designed with one main research question in mind:
\begin{itemize}
\item \textit{RQ1:} What are the inherent differences in perceived readability between programming languages?
\end{itemize}

@@

The industry is left out of the population since they are more likely to have a lot of experience with certain languages which may cause them to be biased towards that language. Furthermore, Höst, Regnell and Whohlin \cite{host2000using} make the case that ``The reason to use students as subjects is often that they are available at universities and they are willing to participate in studies as part of courses they attend.'' which is highly appropriate given that we will be asking our peers and those in lower years who will eventually also be required to perform research tasks. The SE\&M program was chosen due to it being accessible for the researchers. More sophisticated sampling methods such as cluster-sampling or random sampling have not been used due to time-constraints.

@@

The five languages represented in the questionnaire were C++, Erlang, Go, Java and Python. They were chosen for multiple reasons. One being that all the students in the SE\&M course will have taken an introductory Java programming course by now and as such should be familiar with the syntax. The other languages were chosen for similar reasons. Erlang was chosen as it's the first language taught after Java and it uses a different development paradigm. C++ was chosen as it is used heavily in project course during the 4th semester and Python was chosen for similar reasons. Golang was picked as we wanted to see how students would perceive a language that is popular but has not really entered the academic landscape as of yet.

@@

Taking into consideration potential biases that may occur, the code snippets were designed so that the code should be easily understood by first year students who have only had exposure to Java. For example, each code snippet has a standard ``Hello World'' example and a simple for-loop with a counter that is returned at the conclusion of the loop. The only exception to this is Erlang. While Erlang does support for-loops, recursion is the tool of choice to perform iterations. As such, we feel it was a valid example and should not affect the results for this specific language drastically considering, by this point and time in their education, all SE\&M students will have been exposed to recursion. The choices of languages do leave us open to a potential validity and bias issue. Inexperienced students who have only seen Java may rate it as easily readable purely because they are only familiar with Java, this is why the relative level of experience is requested first in the survey, so that we might be able to pick up this potential bias during the results analysis. To help alleviate this issue, the questionnaire was distributed to all three year groups currently studying at SE\&M (2012,2013,2014).

@@

\begin{table}[ht]
  \centering
  \begin{tabularx}
  {\linewidth}{| l | X | X |}
    \hline
    \textbf{Variable Name} & \textbf{Research Question} & \textbf{Item on Survey} \\ \hline
    Experience & RQ1, SQ1 & 1 \\ \hline
    Academic Year & RQ1, SQ1 & 2 \\ \hline
    Preferred Language & RQ1, SQ2 & 3 \\ \hline
    Least Preferred Language & RQ1, SQ2& 4 \\  \hline
    Experience of Language & RQ1, SQ2 & 5 \\ 
    \hline
  \end{tabularx}
  \caption{Survey variables' relation to Research Questions}
\end{table}

@@

\SubSection{Threats to Validity}
As in all research there are threats to the validity of the study being undertaken. The comparison of the perceived readability of programming languages cannot be fully determined from a single study, as such, more are required to make more concrete generalisations. Given this, the threats to internal and construct validity are among the most important for this study. The most import however would be the external validity of the study.

The students were selected from the three current years studying at the University of Gothenburg. Most of the respondents indicated that they are in the first year of their studies and as such their inexperience could be a limiting factor on their answers to the survey. The students may also have had a bias for one of the languages; Java. So far, the students in the first year have been formally taught in only Java. This means that there is a lower chance they would be familiar with reading other language syntax other than Java and as such they would be more likely to see Java favourably than others. We cannot therefore generalise that Java has higher readability than that of the other languages despite what the results show.

A further threat that we were unable to counteract adequately was the willing to participate of our respondents. The participants had little or no incentive to actually partake in the study and as such many most likely simply ignored it. This is to be expected when one attempts to undertakes a single stage sampling method. One idea to improve the response rate for further surveys could be to form some incentive for the respondents.

The fact that the questionnaire was distributed online via Facebook groups also meant that there was another threat to the validity of the responses. These groups are not officially controlled by the university and as such admission into these groups is up to the students who are organising them. This means that there is potential that some of the responses came from people not within the SE\&M institutions and this should be taken into consideration when analysing the results.

Other potential validity threats we must take into consideration are those that arise from the nature of survey. Since we are performing single stage data collection, the respondents of the surveys are answering our questions at their convenience. This opens the door for some validity threats, such as the willingness to participate. Our data are at the mercy of those willing to participate; we cannot anticipate how many will respond. There is little incentive for the respondents to partake in the survey other than for their own self interest.

@@

\vfill
\eject

>>>>

%
%  $Description: Author guidelines and sample document in LaTeX 2.09$ 
%
%  $Author: ienne $
%  $Date: 1995/09/15 15:20:59 $
%  $Revision: 1.4 $
%
\documentclass[10pt,twocolumn]{article} 
\usepackage{latex8}
\usepackage{times}
\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage{tabularx}

@@

%------------------------------------------------------------------------- 
\Section{Introduction}
%Problem and definition
Comparisons between programming languages are difficult to perform, due to inherent differences between languages, such as syntax. One aspect of code is its readability, which in turn is a determining factor of a piece of software's quality and maintainability \cite{aggarwal2002integrated},\cite{elshoff1982improving}. Buse \& Westley define readability as ``a property that influences how easily a given piece of code can be read and understood'' \cite{buse2010learning}.

@@

Industry and academia both face the problem of code maintainability and by association, readability. As was highlighted previously, the readability of code can affect its maintainability. This study could help some in industry to decide which languages should be used if maintainability is of high concern. Further studies could be conducted to determine if there is a correlation between the popularity of projects and the readability of their languages. Furthermore, academia would be interested in which languages are more popular and those that have a higher readability, especially for novice developers. Further investigation into the design of these programming languages could yield insights into what makes them so popular and potentially well suited for teaching.

The purpose of this survey study is therefore to discover how different levels of experience affect the perceived readability of code.

@@

%------------------------------------------------------------------------- 
\SubSection{Research Questions}
This study has been designed with one main research question in mind:
\begin{itemize}
\item \textit{RQ1:} How do different programming languages differ in readability?
\begin{itemize}
\item \textit{SQ1:} How does readability correlate to popularity?
\end{itemize}
\item \textit{RQ2:} How differently do students of differing levels of experience rate code readability?
\end{itemize}

@@

The industry is left out of the population since they are more likely to have a lot of experience with certain languages which may cause them to be biased towards that language. Furthermore, Höst, Regnell and Whohlin \cite{host2000using} make the case that ``The reason to use students as subjects is often that they are available at universities and they are willing to participate in studies as part of courses they attend.'' which is highly appropriate given that we will be asking our peers and those in lower years who will eventually also be required to perform research tasks. The SE\&M program was chosen due to it being accessible for the researchers.

@@

The five languages represented in the questionnaire were C++, Erlang, Go, Java and Python. They were chosen for multiple reasons. One being that all the students in the SE\&M course will have taken an introductory Java programming course by now and as such should be familiar with the syntax. The other languages were chosen for similar reasons. Erlang was chosen as it's the first language taught after Java and it uses a different development paradigm. C++ was chosen as it is used heavily in project course during the 4th semester and Python was chosen for similar reasons. Golang was picked as we wanted to see how students would perceive a language that is popular but has not really entered the academic landscape as of yet. The code snippets do not show which language they are written in. This was done to help eliminate bias towards known languages by their names.

@@

Taking into consideration potential biases that may occur, the code snippets were designed so that the code should be easily understood by first year students who have only had exposure to Java. For example, each code snippet has a standard ``Hello World'' example and a simple for-loop with a counter that is returned at the conclusion of the loop. The only exception to this is Erlang. While Erlang does support for-loops, recursion is the tool of choice to perform iterations. As such, we feel it was a valid example and should not affect the results for this specific language drastically considering, by this point and time in their education, all SE\&M students will have been exposed to recursion. The choices of languages do leave us open to a potential validity and bias issue. Inexperienced students who have only seen Java may rate it as easily readable purely because they are only familiar with Java, this is why the relative level of experience is requested first in the survey, so that we might be able to pick up this potential bias during the results analysis. Complicated code snippets were therefore excluded as we were not specifically aiming for a ``Code reading'' exercise. The questionnaire was distributed to all three year groups currently studying at SE\&M (2012,2013,2014).

@@

\begin{table}[ht]
  \centering
  \begin{tabularx}
  {\linewidth}{| l | X | X |}
    \hline
    \textbf{Variable Name} & \textbf{Research Question} & \textbf{Item on Survey} \\ \hline
    Experience & RQ1 & 1 \\ \hline
    Academic Year & RQ1 & 2 \\ \hline
    Preferred Language & RQ1, SQ1 & 3 \\ \hline
    Least Preferred Language & RQ1, SQ1 & 4 \\  \hline
    Experience of Language & RQ1, SQ1 & 5 \\ 
    \hline
  \end{tabularx}
  \caption{Survey variables' relation to Research Questions}
\end{table}

@@

\SubSection{Threats to Validity}
As in all research there are threats to the validity of the study being undertaken. The comparison of the perceived readability of programming languages cannot be fully determined from a single study, as such, more are required to make more concrete generalisations. Given this, the threats to internal and construct validity are among the most important for this study with the most import being the external validity of the study.

Threats to \textit{external validity} include using students as participants, the size of the code snippets used in the survey and the distribution of the survey. Using students as participants is a threat mainly due to their experience levels. This threat is diminished as we are interesting in the disparity between the levels of experience, furthermore, the use of students in surveys is recommended by Höst, Regnell and Whohlin \cite{host2000using}. The size and complexity of the code snippets provided was minimised. The students have all had at least one course in Java and as such we felt that the simplicity of the code snippets used for the other languages would be appropriate.

As mentioned previously, a pilot study was conducted among peers and students whom we supervise. The goal of this action was to identify and thereby reduce ambiguity in our questionnaire, thus limiting \textit{construction validity} threats.

In regards to \textit{internal validity} threats, we had to cope with performing a convenience survey and random distribution of the surveys. The participants had little or no incentive to actually partake in the study and as such many most likely simply ignored it. One idea to improve the response rate for further surveys could be to form some incentive for the respondents.

The questionnaire being distributed online via Facebook groups was another threat to the \textit{internal validity}. These groups are not officially controlled by the university and as such admission into these groups is up to the students who are organising them. This means that there is potential that some of the responses came from people not within the SE\&M institutions and this should be taken into consideration when analysing the results. One way to overcome this in future studies could be to use formal mailing lists, thus limiting the participants to only those students whom are enrolled at the university.

A final \textit{internal validity} threat is that the students were selected from the three current years studying at the University of Gothenburg. Most of the respondents indicated that they are in the first year of their studies and as such their inexperience could be a limiting factor of their responses. The students may also have had a bias for one of the languages; Java. So far, the students in the first year have been formally taught in only Java. This means that there is a lower chance they would be familiar with reading other language syntax other than Java and as such they would be more likely to see Java favourably than others. This is both an internal and external threat and in order to overcome it, we would most likely need to perform further studies.

@@
